Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-26 paper.tex
mostly spacing for tables / figure, a few typography things too

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple’s M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA’s subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

@@

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

LOBE adds an additional condition on top of OBE: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.

@@

\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained-Propagation} phase entirely.
        \end{itemize}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & LOBE & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40  & 16  \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80  & 32  \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160 & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

@@

\begin{table*}
  \centering
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
    Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
    Avg Lookback Length per Wg  & 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

@@

\subsection{Simulated Blocking}\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins.\label{fig:total_spins}}
  \end{subfigure}

>>>>

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

@@

\subsection{The Scan Primitive}%
\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.

@@

\begin{algorithm}
  \small
  \SetAlgoLined%
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained-Propagation} phase entirely.
        \end{itemize}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
    Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
    Avg Lookback Length per Wg  & 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

@@

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins.\label{fig:total_spins}}
  \end{subfigure}
